---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Speciation Genomics {#ch08}

```{r html-doc, child="html_doc.Rmd", echo = FALSE}
```

```{r setup, include=FALSE}
source("setup.R")
library(tidyverse)
```

```{r, echo = FALSE}
knitr::opts_chunk$set(fig.width = 8)
```

Genomic data has revolutionised the way we conduct speciation research over the past decade. With high-throughput sequencing it is now possible to examine variation at thousands of markers from across the genome. Genome-wide studies of genetic differentiation, particularly measured using *F*~ST~ have been used to identify regions of the genome that might be involved in speciation. The rationale is relatively simple, *F*~ST~ is a measure of genetic differentiation and when species diverge in the presence of gene flow, we might expect that genome regions underlying traits that prevent gene flow between species will show a higher level of *F*~ST~ than those that do not. In other words, genome scan analyses can, in principle, be used to identify barrier loci involved in the speciation process. This approach became extremely popular in many early speciation genomic studies but it overlooked a crucial point - that other processes, not related to speciation can produce the same patterns in the genome. In this session, we will leverage our ability to handle high-throughput, whole genome resequencing data to investigate patterns of nucleotide diversity, genetic differentiation and genetic divergence across a chromosome. We will examine what might explain some of the patterns we observe and learn that while genome scans can be a powerful tool for speciation research, they must be used with caution.

### What to expect {.unnumbered}

In this section we will:

-   Learn tools for visualizing data with many dimensions
-   contrast and compare genome-wide measures of *F*~ST~ among species
-   examine variation in recombination rate and it's influence on differentiation

### Getting started {.unnumbered}

The first thing we need to do is set up the R environment. Today we'll be using `tidyverse` and the `GenoPop` package that we installed and loaded in the [last session](#ch07).

```{r, eval = TRUE, echo = TRUE, results = "hide", message = FALSE, warning = FALSE}
# clear the R environment
rm(list = ls())
library(tidyverse)
library(GenoPop)
```

## Visualizing complex data

```{r html-doc, child="html_doc.Rmd", echo = FALSE}
```

You have previously learned how to use aesthetics in `ggplot` to show many variables in a single plot (e.g., coloring points by group). Today you will learn a bit more about this. We will be working with the [2020 population data](https://bios1140.github.io/data/worlddata.csv) from the very first week, so make sure that is in your working directory.

We start by reading in the data (see if you manage to do this yourself before looking at my code):

::: {.fold .c}
```{r, eval = FALSE}
d <- read.table("worlddata.csv", header = TRUE, sep = ",")
```
:::

```{r, echo = FALSE}
d <- read.table("docs/data/worlddata.csv", header = TRUE, sep = ",")
```

For this tutorial, we will briefly investigate the correlation between fertility rate and life expectancy in the countries of the world. We can start by making a simple scatterplot.

**Exercise: make a scatterplot of fertility rate against life expectancy using the world data**

::: {.fold .s .o}
```{r}
g <- ggplot(d, aes(Total_Fertility_Rate, Life_Expectancy_at_Birth))
# note: adding alpha = 0.5 to better see all points
g + geom_point(alpha = 0.5)
```
:::

This shows a negative correlation between the variables: in countries with lower life expectancy, more children are born per woman. We can investigate this further by dividing the data by continent, and gain some additional perspective by showing the population sizes of each country.

**Exercise: Update the plot by mapping colour to continent, and point size to population size.**

::: {.fold .s .o}
```{r}
h <- g + geom_point(aes(col = Continent, size = Population), alpha = 0.5)
h
```
:::

### Faceting

This is already quite a good plot, showing a lot of data at once. However, it can be a bit difficult to see trends within each continent (if that's what we want to investigate), and what if we want to use colour for something else while still showing differences between continents? One way to do this is to divide the plot into several windows based on a variable, known as **faceting**.

Faceting in `ggplot` can be done by adding the function `facet_wrap()`. The syntax is a bit weird: `facet_wrap(~Variable)`. It uses the `~` (tilde) character[^exercise8-1], and is an exception to the "all variables go inside `aes()`"-rule that we have emphasized earlier. For our plot it looks like this:

[^exercise8-1]: Which can be read as "modeled by" or simply "by", i.e., "facet by `Variable`".

```{r}
h + facet_wrap(~Continent)
```

You can control the layout by using the `nrow` and `ncol` arguments to specify numbers of rows and/or columns:

::: {.fold .o}
```{r}
h + facet_wrap(~Continent, nrow = 3)
h + facet_wrap(~Continent, ncol = 1)
```

Note that all axes are the same across the facets. This can be changed with the argument `scales`, where you can specify "free", "free_y" or "free_x". Be aware that this can be misleading in some cases (like this one, I would argue), so use it with caution! "free_x" is shown below, but try the others yourself to see what happens!

```{r}
h + facet_wrap(~Continent, scales = "free_x")
```
:::

Now you have learned some tools for visualising various statistics across the sparrow genome for later. Let's jump into the evolutionary biology part!

## Returning to the sparrow dataset

```{r html-doc, child="html_doc.Rmd", echo = FALSE}
```

In the last session, we used the `GenoPop` package to calculate sliding window estimates of nucleotide diversity across chromosome 8 of the house sparrow with data from [Ravinet *et al.* (2018)](http://rspb.royalsocietypublishing.org/content/285/1884/20181246). We will now return to this example and use it to demonstrate why we must interpret the genomic landscape of differentiation with caution.

### Preparing to read in the sparrow vcf

The first step we need to take is to read our VCF of the sparrow chromosome 8 into the R environment. This is exactly the same procedure as the [last session](#ch07) but just in case you missed those steps, here they are again.

-   First, download the [VCF](https://bios1140.github.io/data/sparrow_chr8_downsample.vcf.gz) and it's [index](https://bios1140.github.io/data/sparrow_chr8_downsample.vcf.gz.tbi) (we need this to read it in to R)
-   Move the downloaded VCF into your working directory (use `getwd` if you don't know where that is)

We will work with this file soon, but first some housekeeping. We should set a variable to the path of this VCF. Once it is in your working directory:

```{r, eval = FALSE, echo = TRUE, results = 'hide'}
vcf_path = "./sparrow_chr8_downsample.vcf.gz")
```

We eventually want to investigate differences between populations, but the data does not currently contain information about the populations, only individuals. Download the [population data](https://bios1140.github.io/data/sparrow_pops.txt) and put it in your working directory. The following code reads in the population data and creates a vector of individuals which are house sparrows.

::: {.yellow}
```{r, eval = FALSE, results = 'hide'}
sparrow_info <- read.table("sparrow_pops.txt", sep = "\t", header = TRUE)
house <- sparrow_info$ind[sparrow_info$pop == "House"]
```
:::

```{r, eval = TRUE, echo = FALSE, results = 'hide'}
sparrow_info <- read.table("./docs/data/sparrow_pops.txt", sep = "\t", header = TRUE)
house <- sparrow_info$ind[sparrow_info$pop == "House"]
```

You can also create these vectors for each of the other species present. Have a look at the `sparrow_info` object to see what they are. We will also do this below. 

```{r, eval = TRUE, echo = TRUE, results = 'hide'}
bactrianus <- sparrow_info$ind[sparrow_info$pop == "Bactrianus"]
spanish <- sparrow_info$ind[sparrow_info$pop == "Spanish"]
italian <- sparrow_info$ind[sparrow_info$pop == "Italian"]
tree <- sparrow_info$ind[sparrow_info$pop == "Tree"]
```

Now you are done, we are ready to calculate some parameters on genomic data!

## Setting up sliding windows

```{r html-doc, child="html_doc.Rmd", echo = FALSE}
```

So far, this will start to seem quite familiar! We learned in the last session that per-SNP estimates of statistics such as $\pi$ can often be extremely noisy when you are calculating them on very large numbers of markers. As well as this, there are issues with the fact that SNP positions in close proximity are not always independent due to recombination - this is a theme we will return too shortly. So for this reason, it is often better to use a **sliding-window** approach - i.e. split the genome into windows of a particular size and then calculate the mean for a statistic within that window.

We know already that chromosome 8 is 49,693,984 bp long, so we can get an idea of how many sliding windows we would generate by using some R code. We'll set our sliding window to be 100,000 bp wide - or 100 Kb. We will also set a **step** or **jump** for our window of 25,000 bp - or 25Kb.

```{r, eval = TRUE, echo = TRUE, results = 'hidden', message = FALSE}
# set chromosome size
chr8 <- 49693984

# set window size and window jump
window_size <- 100000
window_jump <- 25000
```

We will shortly be using these objects to run our analysis. 

### Calculating sliding window estimates of nucleotide diversity and differentiation

Now that we have set up the R environment, the population information and the sliding windows, it is quite straightforward for us to calculate some statistics we are interested in. In this case, we are going to calculate nucleotide diversity (i.e. $\pi$) and *F*~ST~. We will also generate a third statistic, *d*~XY~, which is the absolute nucleotide divergence between two populations.

First we will calculate $\pi$. Here we will do this for the house sparrow, like you did in the previous exercise. You can also calculate this for the different species too if you want to. 

```{r, eval = FALSE, echo = TRUE, results = 'hide', message = FALSE}
# calculate nucleotide diversity - house
pi_windows_house <- Pi("./sparrow_chr8_downsample.vcf.gz", exclude_ind = c(bactrianus, spanish, italian, tree),
                 seq_length = chr8, window_size = window_size, skip_size = window_jump)
# calculate nucleotide diversity - bactrianus
pi_windows_bac <- Pi("./sparrow_chr8_downsample.vcf.gz", exclude_ind = c(house, spanish, italian, tree),
                 seq_length = chr8, window_size = window_size, skip_size = window_jump)
```

```{r, eval = TRUE, echo = FALSE, results = 'hide', message = FALSE}
# calculate nucleotide diversity - house
pi_windows_house <- Pi("./docs/data/sparrow_chr8_downsample.vcf.gz", exclude_ind = c(bactrianus, spanish, italian, tree),
                 seq_length = chr8, window_size = window_size, skip_size = window_jump)
# calculate nucleotide diversity - bac
pi_windows_bac <- Pi("./docs/data/sparrow_chr8_downsample.vcf.gz", exclude_ind = c(house, spanish, italian, tree),
                 seq_length = chr8, window_size = window_size, skip_size = window_jump)
```

While we wait we can break this down:
  - `vcf_file` - just defines the path to our vcf file
  - `exclude_ind` - sets individuals to leave out of the analysis (i..e everything but house here)
  - `seq_length` - this is the length of the chromosome, taken from the genome information
  - `window_size` is the size of the windows we examine, here 100Kb
  - `skip_size` is the step size for our sliding windows - 25Kb in this instance. 

Next we will calculate *F*~ST~, which again is very straight forward with a single command. Here we will do this for house versus bactrianus sparrows. Feel free to experiment with other comparisons if you wish. 

```{r, eval = FALSE, echo = TRUE, results = 'hide', message = FALSE}
# calculate Fst
fst_windows_house_bac <- Fst(vcf_path = vcf_path, pop1_individuals = house, pop2_individuals = bactrianus,
                             weighted = TRUE, window_size = window_size, skip_size = window_jump)
```

```{r, eval = TRUE, echo = FALSE, results = 'hide', message = FALSE}
# calculate Fst
fst_windows_house_bac <- Fst("./docs/data/sparrow_chr8_downsample.vcf.gz", pop1_individuals = house, pop2_individuals = bactrianus,
                             weighted = TRUE, window_size = window_size, skip_size = window_jump)

```

Note that here we use `weighted = TRUE` to specify we want the weighted *F*~ST~ per window. This just means that our *F*~ST~ estimates are weighted by the number of SNPs in each window, such that windows with many versus those with very few SNPs are accounted for . 

Finally, we will also repeat this for *d*~XY~, which is equally straightforward.

```{r, eval = FALSE, echo = TRUE, results = 'hide', message = FALSE}
# calculate Dxy
dxy_windows_house_bac <- Dxy(vcf_path = vcf_path, pop1_individuals = house, pop2_individuals = bactrianus,
                             seq_length = chr8, window_size = window_size, skip_size = window_jump)
```

```{r, eval = TRUE, echo = FALSE, results = 'hide', message = FALSE}
# calculate Dxy
dxy_windows_house_bac <- Dxy(vcf_path = "./docs/data/sparrow_chr8_downsample.vcf.gz", pop1_individuals = house, pop2_individuals = bactrianus,
                             seq_length = chr8, window_size = window_size, skip_size = window_jump)

```

The only difference here is that we again need to specify the length of the chromosome with `seq_length = chr8` - i.e. the variable we set up earlier. 
And that's it for calculating the statistics! 

### Extracting statistics for visualisation {#sparrow-data}

Since we ran our analysis on a sliding-window basis, we should have estimates of $\pi$, *F*~ST~ and *d*~XY~ for each window. What we want to do now is extract all our statistics and place them in a single `data.frame` for easier downstream visualisation - this will let us identify how these statistics are interrelated.

First up, we will combine all our statistics into a single data.frame, sort out the classes of some of the data and then make it a tibble for easier access. One issue we have to deal with is that the analysis produces `NA` for windows with no SNPs - this means the values of $\pi$ differ for the house and bactrianus analyses. Below we first merge the $\pi$ datasets to deal with this and then create our master dataset for visualisation. The last step just adds a midpoint value for all our windows for plotting later on. 

::: {.yellow}
```{r, eval = TRUE, echo = TRUE, results = 'hidden', message = FALSE}
# Join the pi datasets
pi <- left_join(pi_windows_house, pi_windows_bac, by = "Start", suffix = c("_house", "_bac"))

# create the master dataset 
sparrow_data <- data.frame(
  chromosome = as.character(pi_windows_house$Chromosome),
  start =  as.numeric(pi_windows_house$Start),
  end = as.numeric(pi_windows_house$End),
  pi_house = as.numeric(pi$Pi_house),
  pi_bac = as.numeric(pi$Pi_bac),
  fst = as.numeric(fst_windows_house_bac$Fst),
  dxy = as.numeric(dxy_windows_house_bac$Dxy)
) %>% as_tibble()

# get midpoint of genome windows
sparrow_data <- sparrow_data %>%
  mutate(mid = round((start + end)/2))
```
:::

We now have the data frame `sparrow_data`, take a look at it to ensure that it looks correct (be aware of any errors you might get while running the code in yellow above). The data contain window positions as well as nucleotide diversity $\pi$ and the two pairwise measures F~ST~ and d~xy~ for our house and bactrianus comparison. Now we can finally investigate differences between the populations!

## Visualising the data {#sparrow-viz}

```{r html-doc, child="html_doc.Rmd", echo = FALSE}
```

For the purposes of this session, we will focus mainly on the difference between house and bactrianus sparrows. Now have all our data in a tidy `data.frame`, it is very easy to calculate things like the mean values of our statistics among all the different species. For example, let's say we want to look at mean nucleotide diversity, we can do that like so:

```{r, eval = FALSE, echo = TRUE, results = 'hidden', message = FALSE}
# select nucleotide diversity data and calculate means
sparrow_data %>% select(contains("pi")) %>% summarise_all(mean, na.rm = T)
```

A lot of this will be familiar from before but to clarify, we used `select()` and `contains()` to select columns from our main dataset that contain the string `"pi"`---i.e. nucleotide diversity columns. We then used `summarise_all` and `mean` to calculate the mean value for all of the columns we selected. The `na.rm = T` part is just a call to the `mean` function to ignore any `NA` values. From the output above, we can see that the house sparrow has the highest levels of nucleotide diversity.

We could also quite easily plot if we wanted to. However, to do this, we need to use `pivot_longer` on the data to get all species in one column, and all $\pi$ values in another[^exercise8-4]. Note how we can supply `contains("pi")` inside `pivot_longer()` to specify which columns to include.

[^exercise8-4]: It may have been a while since the last time you saw `pivot_longer()`. If you have forgotten how it works or why we use it, remember that you can always go back to [Week 2](#pivot-longer).

```{r, eval = TRUE, echo = TRUE, results = 'hidden', message = FALSE}
# gather the data
pi_g <- sparrow_data %>%
  pivot_longer(contains("pi"), names_to = "species", values_to = "pi")
# make a boxplot
a <- ggplot(pi_g, aes(species, pi)) + geom_boxplot() + theme_light() + xlab(NULL)
a
```

This makes it much clearer how nucleotide diversity differs between the two lineages.

### Visualising patterns along the chromosome

Let's have a look at how *F*~ST~ between house and bactrianus sparrows varies along chromosome 8. Note that we plot the mid-point of each window, and divide the position by $10^6$ to get megabases on the x-axis.

```{r, eval = TRUE, echo = TRUE, results = 'hidden', message = FALSE}
a <- ggplot(sparrow_data, aes(mid/10^6, fst)) + geom_line(colour = "purple")
a <- a + xlab("Position (Mb)") + ylab(expression(italic(F)[ST]))
a + theme_light()
```

From this plot, it is clear there is a huge peak in *F*~ST~ around 30 Mb. Actually, there are several large peaks on this genome but is this one a potential region that might harbour a speciation gene? Well you might recall from the previous session that there is a drop in nucleotide diversity in this region...

How can we investigate this? The easiest thing to do is to plot $\pi$, *F*~ST~ and *d*~XY~ to examine how they co-vary along the genome. This requires a bit of data manipulation, which we will break it down into steps.

First, let's get the data we are interested in:

```{r, eval = TRUE, echo = TRUE, results = 'hidden', message = FALSE}
# select data of interest
hs <- sparrow_data %>% select(mid, pi_house, pi_bac, fst, dxy)
```

To keep things simple, we've thrown everything out we don't need. Next, we need to use `pivot_longer` in order to rearrange our `data.frame` so that we can plot it properly.

```{r, eval = TRUE, echo = TRUE, results = 'hidden', message = FALSE}
# use pivot_longer to rearrange everything
hs_g <- pivot_longer(hs, -mid, names_to = "stat", values_to = "value")
```

Here, we use `-mid` to tell the function we want to leave this out of the pivoting and use `names_to = "stat"` to make it clear we are arranging our data by the statistics we have calculated, `values_to = "value"` is just a name for the values of each of our statistics.

Now we can plot everything together:

```{r, eval = TRUE, echo = TRUE, results = 'hidden', message = FALSE}
a <- ggplot(hs_g, aes(mid/10^6, value, colour = stat)) + geom_line()
a <- a + xlab("Position (Mb)")
a + theme_light()
```

OK so it should be immediately obvious that this plot is really unhelpful. We see the *F*~ST~ data again, but since that is on such a different scale to estimates of $\pi$ and *d*~XY~, we can't see anything! Instead, it would make a lot more sense to split our plot into facets - i.e. a plot panel for each statistic. Lucky for us, we learned to facet plots with `facet_grid` in the beginning of this tutorial! Remember that we can specify independent y-axes with `scales = "free_y"`, and set `ncol = 1` to get all plots below each other.[^exercise8-5]



```{r, eval = TRUE, echo = TRUE, results = 'hidden', message = FALSE}
# construct a plot with facets
a <- ggplot(hs_g, aes(mid/10^6, value, colour = stat)) + geom_line()
a <- a + facet_wrap(~stat, scales = "free_y", ncol = 1)
a <- a + xlab("Position (Mb)")
a + theme_light() + theme(legend.position = "none")
```

[^exercise8-5]:

    In this case, the plot might be easier to interpret if we rearranged everything so *F*~ST~ came at the top, $\pi$ beneath it and then finally, *d*\_XY\_. We can use the function `fct_relevel()` for manually reordering the factors to achieve this:
    
    ```{r, eval = TRUE, echo = TRUE, results = 'hidden', message = FALSE}
    new_order <- c("fst", "pi_house", "bac_pi", "dxy")
    hs_g$stat <- fct_relevel(hs_g$stat, new_order)
    ```
    
    We can now replot our figure with the new order:
    
    ```{r, eval = TRUE, echo = TRUE, results = 'hidden', message = FALSE}
    # construct a plot with facets
    a <- ggplot(hs_g, aes(mid/10^6, value, colour = stat)) + geom_line()
    a <- a + facet_wrap(~stat, scales = "free_y", ncol = 1)
    a <- a + xlab("Position (Mb)")
    a + theme_light() + theme(legend.position = "none")
    ```

Examining the plot we created, it is pretty clear that the large peak in *F*~ST~ on our chromosome is matched by two regions of low nucleotide diversity in the house and bactrianius sparrow, *d*~XY~ is also very low in the same region.

The signal here is quite clear - what could explain it? Low recombination regions in the genome are one potential explanation. The reason for this is that in a low recombination region, **background selection** and also **selective sweeps** can remove variation at polymorphic positions that are closely linked to the target of selection. Selection of this kind in either the house or the bactrianus lineages AFTER they have split into different species will reduce $\pi$ in these regions and since *F*~ST~ is a relative measure of differentiation, it will potentially be inflated.

This is an important issue as it means that we cannot reliably use *F*~ST~ to identify genes involved in reproductive isolation or possibly even local adaptation from a genome scan. By comparing *F*~ST~ to *d*~XY~ here, we see the latter is also reduced in this region, which again suggests it is likely that some sort of genome structure might be responsible for the peak in *F*~ST~ we see. One way to investigate this is examine the recombination rate variation along chromosome 8---which we will do next, after a short interlude.

::: {.green}
### Interlude: relative vs. absolute measures of nucleotide diversity

We have mentioned a couple of times that *F*~ST~ is a *relative* measure of nucleotide differences, while *d*~xy~ is an *absolute* measure. But what does this mean? According to [Wikipedia](https://en.wikipedia.org/wiki/Fixation_index#Estimation), a simple way of estimating F~ST~ (there are many ways to do this) is using the formula:

$$ F_{ST} = \frac{\pi_{between} - \pi_{within}}{\pi_{between}} $$

Thus, *F*~ST~ takes into account not only the nucleotide diversity *between* the populations, but also the diversity *within* the populations. So, what if the diversity within is very low, for example 0? Putting that into the formula above, we see that no matter what $\pi_{between}$ is, *F*~ST~ will be 1.

In other words, *F*~ST~ is sensitive to the variation within each population, which is something you should be aware of before drawing conclusions. *d*~xy~ does not take the diversity within into account, and is thus an *absolute* measure of differences.
:::

### Investigating recombination rate variation

To check whether variation in recombination might explain the pattern we observed, we will read in the recombination rate estimated for 100 Kb windows with a 25 Kb step on chromosme 8. This was originally estimated from a house sparrow linkage map, published by [Elgvin et al (2018)](http://advances.sciencemag.org/content/3/6/e1602996) and you can download the data [here](https://bios1140.github.io/data/chr8_recomb.tsv). We will read the data in like normal

```{r, eval = FALSE, echo = TRUE, results = 'hidden', message = FALSE}
rrate <- read.table("chr8_recomb.tsv", sep = "\t", header = TRUE)
```

```{r, eval = TRUE, echo = FALSE, results = 'hidden', message = FALSE}
rrate <- read_delim("docs/data/chr8_recomb.tsv", delim = "\t")
```

Unfortunately, the recombination rate data has more rows than our dataset. So we need to join them together. 

```{r, eval = TRUE, echo = TRUE, results = 'hidden', message = FALSE}
# First make a tibble
rrate <- as_tibble(rrate)
# ensure midpoints match
rrate <- rrate %>% mutate(mid = mid - 1)
# match to the sparrow_data
sparrow_data <- left_join(sparrow_data, rrate %>% select(-chr), by = "mid")
```

Now we are ready to see whether the variation in nucleotide diversity and *F*~ST~ can be explained by recombination rate. Let's plot how it varies along the genome.

```{r, eval = TRUE, echo = TRUE, results = 'hidden', message = FALSE}
# construct a plot for recombination rate
a <- ggplot(sparrow_data, aes(mid/10^6, recomb)) + geom_line()
a <- a + xlab("Position (Mb)") + ylab("Recombination rate (cM/Mb)")
a + theme_light() 
```

To explain this a little, we have plotted recombination rate in **centiMorgans per Megabase** - i.e. essentially the probability that a recombination event can occur. The higher this value is, the higher the probability of recombination. The first obvious point to take home from this figure is that our recombination rate varies quite significantly across the genome. Secondly, we see quite a drastic reduction in recombination rate between about 23 Mb and 30 Mb. This is exactly where our *F*~ST~ peak occurs. to confirm this, we will plot both statistics together.

```{r, eval = TRUE, echo = TRUE, results = 'hidden', message = FALSE}
# subset data and gather
hr <- sparrow_data %>% 
  select(mid, fst, recomb) %>%
  pivot_longer(-mid, names_to = "stat", values_to = "value")
# make a facet plot
a <- ggplot(hr, aes(mid/10^6, value)) + geom_line()
a <- a + facet_wrap(~stat, scales = "free_y", ncol = 1)
a <- a + xlab("Position (Mb)") + ylab("Recombination rate (cM/Mb)")
a + theme_light() 
```

When we plot our data like this, it is actually more clear that perhaps both of the large peaks on chromosome 8 occur in an area of very low recombination. What could be causing such low recombination? Well one possibility is the [centromere](https://en.wikipedia.org/wiki/Centromere) is likely to be present here.

Now that we have recombination data read into R, we can also explore the relationships between recombination rate and other statistics in more detail. To demonstrate this, we will plot the joint distribution of recombination rate and *F*~ST~ between house and bactrianus sparrows.

```{r, eval = TRUE, echo = TRUE, results = 'hidden', message = FALSE}
# plot recombination rate and fst
a <- ggplot(sparrow_data, aes(recomb, fst)) + geom_point()
a <- a + xlab("Recombination rate (cM/Mb)") + ylab(expression(italic(F[ST])))
a + theme_light() 
```

Clearly there is a bias here - higher *F*~ST~ values are found in regions of low recombination. Although this doesn't completely invalidate the use of *F*~ST~ in speciation genomics, it does mean we must be cautious when using it to identify genes involved in speciation. If we had not done so here, it would have been quite easy to mistake the peak on chromosome 8 as having an important role in maintaining reproductive isolation between house and bactrianus sparrows.

## Study questions

The study questions for week 6-7 can be found [here](#w07). Deliver them in Canvas before the deadline after week 7 as a word or pdf document. See [the appendix](#rmarkdown) for some important points on how the assignments should be delivered. There, you will also find an introduction to R Markdown, a good way to combine code, output and text for a report.

## Going further

-   [A nice review on what we mean by speciation genomics](https://academic.oup.com/biolinnean/article/124/4/561/5035934)
-   [Some more information on performing whole-genome analyses in PopGenome](https://cran.r-project.org/web/packages/PopGenome/vignettes/Whole_genome_analyses_using_VCF_files.pdf)
-   [Additional learning resources on speciation genomics using Python, R and Unix](http://evomics.org/learning/population-and-speciation-genomics/2018-population-and-speciation-genomics/)
