---
title: "The  Theory of Natural Selection"
output: 
  bookdown::html_document2:
    toc: true
    toc_float: true
    highlight: pygments
    css: "style.css"
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
source("setup.R")
```

```{r html-doc, child="html_doc.Rmd", echo = FALSE}
#include html script
```


::: {.green}
**Important concept:**  
_Green boxes summarises important concepts from the text._
:::

::: {.blue}
**Additional info**  
_Blue boxes contain small tips and additional information for those that are interested. What's in these is hopefully useful, but not mandatory. Feel free to skip these if you want._
:::

:::{.yellow}
**Advanced code**  
_Yellow boxes contain code that you need to run in order to complete the tutorial, but that you don't necessarily need to understand._
:::


**Text in bold contains small exercises to do on your own throughout the tutorial. These are for your own understanding only, so you don't need to hand them in.**