--- 
title: "R tutorials in BIOS1140"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
link-citations: yes
description: "This is a set of R tutorials connected to the course BIOS1140 at the University of Oslo, and are meant as a companion to the book Evolutionary Genetics (Ravinet & Sætre, 2018). They do not assume any prior R knowledge."
---

# Preface {-}

```{r setup, include=FALSE}
source("setup.R")
library(tidyverse)
```

```{r html-doc, child="html_doc.Rmd", echo = FALSE}
```



Welcome!

::: {.green}
**Important concept:**\
*Green boxes summarises important concepts from the text.*
:::

::: {.blue}
**Additional info**\
*Blue boxes contain small tips and additional information for those that are interested. What's in these is hopefully useful, but not mandatory. Feel free to skip these if you want.*
:::

::: {.yellow}
**Advanced code**\
*Yellow boxes contain code that you need to run in order to complete the tutorial, but that you don't necessarily need to understand.*
:::

**Text in bold contains small exercises to do on your own throughout the tutorials. These are for your own understanding only, so you don't need to hand them in.**
